---
title: "Lecture 3"
author: "Zhentao Shi"
date: "Feb 20, 2017"
output:
  pdf_document: default
  html_document: default
---


## Graphics

An English cliche says "One picture is worth ten thousand words".
John Tukey, a renowned mathematical statistician, was one of the pioneers of statistical graphs
in the computer era. Nowadays, powerful software is able to produce dazzling statistical graphs,
sometimes web-based and interactive. Outside of academia, journalism hooks a wide readership with
professional data-based graphs. New York Times and The Economists are first-rate examples;  
South China Morning Post sometimes also does a respectable job.
A well designed statistical graph can deliver an intuitive and powerful message.
I consider graph prior to table when writing a research report or an academic paper.
Graph is lively and engaging. Table is tedious and boring.

We have seen an example of R graph in the OLS regression linear example in Lecture 1.
`plot` is a generic command for graphs, and can be viewed as the default graph engine.
It is capable of producing preliminary statistical graphs.

Over the years, developers all over the world have had many proposals for
more sophisticated statistical graphs. `ggplot2`, contributed by [Hadley Wickham](http://had.co.nz/),
is the best from my point of view.

Ggplot2 is a sophisticated graphic system that generates high-quality statistical graphs.
It is not possible to cover it in a lecture. Fortunately, the author wrote a good reference
[ggplot2 book](http://link.springer.com/book/10.1007%2F978-0-387-98141-3),
which can be downloaded via CUHK network.

Wickham also developed `reshape2`, a package dedicated mainly to prepare data frames for ggplot2.

The workflow of ggplot is to add the elements in a graph one by one, and then print out
the graph all together.
In contrast, `plot` draws the main graph at first, and then add the supplementary elements later.

Ggplot is particularly good at drawing multiple graphs, either of the same pattern or of
different patterns. Multiple subgraphs convey rich information.


**Example**

Plot the density of two estimators under three different data generating processes.
This is an example to generate subgraphs of the same pattern.

```{r,cache=TRUE}
load("big150.Rdata")
library(ggplot2)
library(reshape2)

big150_1 = big150[, c("typb", "numb",  "b1", "b1_c")]
print(head(big150_1))

big150_1 = melt(big150_1, id.vars = c("typb", "numb"), measure.vars = c("b1", "b1_c"))
names(big150_1)[3] = c("estimator")
print(head(big150_1))

p1 = ggplot(big150_1)
p1 = p1 + geom_area(stat = "density", alpha = .25,
                    aes(x = value, fill = estimator),  position = "identity")
p1 = p1 + facet_grid( typb ~ .  )  # this dataset has numb = 150, but no other sample size
p1 = p1 + geom_vline(xintercept = 0)
p1 = p1 + theme_bw()
p1 = p1 + theme(strip.text = element_text( size = 12),
                axis.text = element_text( size = 12))
print(p1)
```
The function `ggplot` specifies which dataset to use for the graph. `geom_***` determines the shape to draw, either scatter dots, lines, curves or areas. `theme` is to fine tune the supplementary elements like the background, the size and font of the axis text and so on.


**Example**

This example aligns two graphs of different patterns in one page.

```{r,cache=TRUE}
# graph packages
library(lattice)
library(ggplot2)
library(reshape2)
library(gridExtra)

load("multigraph.Rdata") # load data


# unify the theme in the two graphs
theme1 = theme_bw() + theme(axis.title.x = element_blank(),
                            strip.text = element_text( size = 12),
                            axis.text = element_text( size = 12),
                            legend.position = "bottom", legend.title = element_blank())

# sub-graph 1
p1 = qplot( x = 1:480, y = m_vec, geom = "line")
p1 = p1 + theme1 + ylab("fraction of chartists")


# sug-graph 2
d2$month = 1:480
p2 = ggplot(d2)
p2 = p2 + geom_line(  aes(x = month, y = value, col = variable) )  
p2 = p2 + theme1 +  ylab("price and fundamental")

# generate the grahp
grid.arrange(p1, p2, nrow=2)
```

In order to unify the theme of the two distinctive subgraphs,
we define an object `theme1` and apply it in both graphic objects
`p1` and `p2`.
