---
title: "Graphics"
author: "Zhentao Shi"
date: "Feb 7, 2018"
output:
  pdf_document: default
  html_document: default
---


## Graphics

An English cliche says "One picture is worth ten thousand words".
John Tukey, a renowned mathematical statistician, was one of the pioneers of statistical graphs
in the computer era. Nowadays, powerful software is able to produce dazzling statistical graphs,
sometimes web-based and interactive. Outside of academia, journalism hooks a wide readership with
professional data-based graphs. New York Times and The Economists are first-rate examples;
South China Morning Post sometimes also does a respectable job.
A well designed statistical graph can deliver an intuitive and powerful message.
I consider graph prior to table when writing a research report or an academic paper.
Graph is lively and engaging. Table is tedious and boring.

We have seen an example of R graph in the OLS regression linear example in Lecture 1.
`plot` is a generic command for graphs, and is the default R graphic engine.
It is capable of producing preliminary statistical graphs.

Over the years, developers all over the world have had many proposals for
more sophisticated statistical graphs. In my opinion, `ggplot2`, contributed by [Hadley Wickham](http://had.co.nz/),
is the best.

`ggplot2` is an advanced graphic system that generates high-quality statistical graphs.
It is not possible to cover it in a lecture. Fortunately, the author wrote a comprehensive reference
[**ggplot2 book**](http://link.springer.com/book/10.1007%2F978-0-387-98141-3),
which can be downloaded via the CUHK campus network (VPN needed).

Wickham also developed `reshape2`, a package dedicated mainly to prepare data frames for `ggplot2`.

The workflow of ggplot is to add the elements in a graph one by one, and then print out
the graph all together.
In contrast, `plot` draws the main graph at first, and then adds the supplementary elements later.

`ggplot2` is particularly good at drawing multiple graphs, either of the same pattern or of
different patterns. Multiple subgraphs convey rich information and easy comparison.


**Example**

Plot the density of two estimators under three different data generating processes.
This is an example to generate subgraphs of the same pattern.

```{r,cache=TRUE}
load("big150.Rdata")
library(ggplot2)
library(reshape2)

big150_1 = big150[, c("typb", "b1", "b1_c")]
print(head(big150_1))

big150_1 = melt(big150_1, id.vars = "typb", measure.vars = c("b1", "b1_c"))
names(big150_1)[2] = c("estimator")
print(head(big150_1))

p1 = ggplot(big150_1)
p1 = p1 + geom_area(stat = "density", alpha = .25,
                    aes(x = value, fill = estimator),  position = "identity")
p1 = p1 + facet_grid( typb ~ .  )  
p1 = p1 + geom_vline(xintercept = 0)
p1 = p1 + theme_bw()
p1 = p1 + theme(strip.text = element_text( size = 12),
                axis.text = element_text( size = 12))
print(p1)
```
The function `ggplot` specifies which dataset to use for the graph. `geom_***` determines the shape to draw, for example scatter dots, lines, curves or areas. `theme` is to tune the supplementary elements like the background, the size and font of the axis text and so on.


**Example**

This example aligns two graphs of different patterns in one page.

```{r,cache=TRUE}
# graph packages
library(lattice)
library(ggplot2)
library(reshape2)
library(gridExtra)

load("multigraph.Rdata") # load data


# unify the theme in the two graphs
theme1 = theme_bw() + theme(axis.title.x = element_blank(),
                            strip.text = element_text( size = 12),
                            axis.text = element_text( size = 12),
                            legend.position = "bottom", legend.title = element_blank())

# sub-graph 1
d1 = data.frame(month = 1:480, m = m_vec)
p1 = qplot( x = month, y = m, data = d1, geom = "line")
p1 = p1 + theme1 + ylab("fraction of chartists")


# sug-graph 2
d2$month = 1:480
p2 = ggplot(d2)
p2 = p2 + geom_line(  aes(x = month, y = value, col = variable) )  
p2 = p2 + theme1 +  ylab("price and fundamental")

# generate the grahp
grid.arrange(p1, p2, nrow=2)
```

In order to unify the theme of the two distinctive subgraphs,
we define an object `theme1` and apply it in both graphic objects
`p1` and `p2`.
